Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-09-02 Paper/Paper.tex

Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition. The daily tasks a physician faces change drastically when a new system is put in place. Some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards lower cost work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.

@@

It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. First, I consider physicians already of retirement age when a new technology is implemented. While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life. Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), suggesting a negative impact.

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management. While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how EHRs impact job satisfaction of current clinical work. As I discuss in the Data Section, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

@@

For most physicians, EHRs will not impose a costly enough burden to induce retirement. However, there are other ways physicians might change their behavior under sufficient incentive to avoid the technology. For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over the technologies used in the work environment, affecting overall job satisfaction. Generally, physicians may have the ability to move hospitals, move from a hospital to an office, or work in multiple facilities and shift the composition of patients towards other facilities. I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists (for a detailed discussion of included physicians, see Appendix \ref{app:data}). Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.

First, I consider those who are in a relatively binding contract. For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital. This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings. A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set. Since these hospitalists are established in their hospital, they work a significant amount of time there. When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings. Alternatively, there are hospitalists that work in multiple hospitals or work in both hospitals and offices. If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction.

@@

There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation. A lack of response could be due to 3 factors: EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment. For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment. An important aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments. Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.

@@

There is still a large portion of physicians not induced to retire due to EHR exposure. Thus, I consider the decision to switch work settings using multiple outcomes. First, I investigate the extent to which hospitalists work in office and whether this changes due to EHR exposure. I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office in a given year. Second, I investigate whether EHR exposure causes a change in the likelihood of switching zip codes. This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.

I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}. The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .014 percentage points, a 4.6\% increase relative to the mean. This result indicates that hospitalists are more likely to start working in an office in the year they are exposed to an EHR, though the effect is not as large as the one found for physicians retiring. Splitting the sample by age reveals that this result is driven by the younger sample of physicians. There is no evidence that older physicians are shifting towards offices. This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more stuck to their particular work environment. While there is not strong evidence for the existence of a pre-trend, I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}. This analysis yields the same result.

@@

It may be that hospitalists who are already working in office settings shift more of their time or patients towards the office and away from hospitals if EHRs impose a meaningful cost to working in a hospital. Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings. I show the estimates and 95\% confidence intervals for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings. This result has varying implications depending on whether total number of patients (the denominator of this variable) is affected. If total number of patients remains constant, then the absolute number of patients seen in offices is also unchanged. However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in the same direction. I revisit the implications of this result in Section \ref{sec:patientcount}, where I examine the total number of patients as an outcome variable.

\begin{figure}[ht]
    \centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
    \includegraphics[scale=.6]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0003 (0.004), -0.0007 (0.005), -0.006 (0.018), respectively.}}
\end{figure}

The previous results only capture switching behavior which involves office-based settings. Another way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors. Again, I exclude physicians who ever retire. The outcome variable is equal to 1 if their list of primary zip codes changes. This variable is limited in the way it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes. While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation.

The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}. For the full sample of hospitalists, there is a statistically positive effect observed in each year after EHR exposure, but there is also evidence of a pre-trend that may be driving this. To make a more conclusive argument about the true effect under violations of the parallel trends assumption, I include a plot of confidence intervals that are robust to specified violations of parallel trends for each age group in Appendix \ref{sec:pretrends}. These results reveal that under linear violations of the parallel trends assumption, there is still a positive effect of EHR exposure on changing zip codes for younger physicians and no effect for retirement age physicians. A .043 ppt increase in the likelihood of changing zip codes for younger physicians indicates a 36\% increase relative to the mean, a significant increase in probability.

@@

A limitation in this analysis is that physicians may shift work behavior due to the organization of the hospital system changing. For example, a hospital may acquire a smaller practice and send physicians to work there for a period of time. If this acquisition is correlated with incentives to adopt an EHR, then the effect I find is biased upward.

\subsection{Patient Count and Billable Activity}\label{sec:patientcount}

For both patient and claim count, if the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR. Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample. These hospitalists do not retire or see patients in another hospital. Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the sample.

The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}. For hospitalists of any age, patient count increases by 17.5 patients in the year of EHR exposure, 34.7 patients in the year after exposure, 16.6 patients in the second year after exposure, and then no effect after the second year. These results translate to a 5-10\% increase in the number of patients seen relative to the mean. As with other outcomes, this result is totally due to changes by the younger group of physicians. Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends}.

@@

Patient count is also driven by demand for health care. A major change in the U.S. health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook. If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the results. The ACA has potentially opposing effects on Medicare patients: Medicare patients may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population. A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results. Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}. Each group shows the same pattern of a small increase in patient count in the first 0-2 years after implementation.

@@

Since I found in the earlier stages of analysis that physicians may have shifted away from hospitals due to EHR exposure, I also investigate whether this increase in the number of patients seen is due to the decrease in working physicians. For example, if a hospital loses a portion of their physicians, they may divide that physician's patients between the remaining physicians. In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away my results could be driven by the retiring physicians instead of EHR use. In Figure \ref{fig:numphys}, I present a graph showing the relationship between EHR adoption and the number of physicians working in the hospital. While these are not causal estimates of the effect of EHR implementation on the number of physicians working in a hospital, they do show that the trend seems to be more physicians working in hospitals, and there is no apparent shock in the short run after EHR implementation. This suggests that the increase in patient count is not due to the exit of physicians observed in the earlier analysis.

@@

Next, I consider whether EHR exposure affects claim count. EHRs may affect physician incentives in such a way that billing activity changes differently from patient count. If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the number of patients has changed or not. On average for the entire sample, claim counts are roughly 4 times the number of patients seen in a year. Shown in Figure \ref{fig:claim}, there is not strong evidence for an effect of EHR exposure on claim count for physicians of retirement age. The point estimates indicate a positive effect in the first two years of exposure for younger physicians, but this is not strong evidence. Thus, for hospitalists $>= 60$, it seems EHRs are not making significant changes to practice behavior. However, for those $< 60$, the observed claim count increase is two times the observed patient count increase. Since the average claim count per patient is 4, these results imply that the number of claims per patient is actually decreasing. That is, physicians are billing for less items as they increase the number of patients.

\begin{figure}[ht]
    \centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Claim Count}
    \includegraphics[scale=.6]{Objects/claim_plot.pdf}
    \label{fig:claim}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 4.04 (33.89), 25.77 (31.26), -161.06 (133.35), respectively.}}
\end{figure}

@@

In the observed results above, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure. Almost all specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure. One reason for this may be the composition of the hospitals which adopt at different times. Consider event time $t+4$. The physician included in the estimation for this coefficient are those exposed in 2010 compared with those who did not adopt until 2015, implying that the underlying hospitals implemented EHRs in either 2010 or 2015. It is likely that the group of hospitals who adopted in 2015 are compositionally different than early adopters, especially if poorer hospitals adopted later. For this reason, I have focused discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.

@@

One concern with the presented analysis, particularly for the productivity outcomes, is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals. The existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology. Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals. they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator. I will refer to any employee in these categories as a data assistant. In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.

@@

In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes. This technology has been controversial and debated as a burdensome load on physicians that may contribute to physician burnout, and therefore patient outcomes. The interaction between health information technology and labor markets is vastly important as the technology is continually legislated and growing each year.

I utilize data on shared patients, hospital EHR use, and physician claim activity. Using an event study framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR in a hospital, where those of retirement age have an even larger increase relative to younger physicians. Further, I find evidence that younger physicians are switching work settings, but that this may be driven by hospital incentives instead of physician incentives. Finally, physicians who remain in the labor force increase both number of patients seen and billing activity, but billing activity per patient decreases with EHR exposure.

The implications of these results are threefold. First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care. In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss may have long term negative effects on patient outcomes, but this is a question left to future research. Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid burdens externally placed on them. This has implications when considering the current policy movement towards regulated care, that physicians may go to extreme measures to avoid such regulation. Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users. While the technology may be a burden, physicians are able to see more patients because of their use, and do not seem to over-bill because of them.

@@

For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}}. I use data that captures shared patient activity in 30 day intervals. I merge the shared patient data to the filtered tax code data created in Section \ref{sec:taxcode} to identify NPIs who are either PCPs or hospitals. I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs. Some pairs are duplicates from the hospital being listed first or second in the shared patient data. I combine duplicates into one observation, summing the same day count variable. Once duplicated are removed, there are 7.1 million observations. Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital. I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data. The final list of pairs consists of 1.3 million observations.

@@

I merge the remaining pairs to an additional survey, the AHA IT Supplement, only sent to hospitals who reported using an EHR. There are more detailed questions in this survey including the EHR vendor, and the ranking of documentation and decision making features the hospital uses. There is a lot of missing data in this survey, which is why I do not use it in my main analysis.

>>>>

Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition. The daily tasks a physician faces change drastically when a new system is put in place. Some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards lower cost work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Further, the cost imposed is likely to vary based on the type of worker, in particular, age group. Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.

@@

It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. First, I consider physicians already of retirement age when a new technology is implemented. While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), suggesting a negative impact.

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management. While it is unlikely that physicians under some age threshold are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction of current clinical work. As I discuss in the Data Section, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

@@

For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of being in the labor market. However, there are other ways physicians might change their behavior under drastic changes to their current working environment. For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over which technologies they use. Generally, physicians may have the ability to change hospitals, move from a hospital to an office, or work in multiple facilities and change the amount of work done in different facilities. I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists (for a detailed discussion of included physicians, see Appendix \ref{app:data}). Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both. While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faces by each type.

First, I consider those who are in a relatively binding contract with the hospital. For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital. This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings. A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set. Since these hospitalists are established in their hospital, they work a significant amount of time there. When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings. Thus, I classify these physicians as "sticky", since they are not able to change behavior in the short run.

Alternatively, there may be hospitalists that work in multiple hospitals or work in both hospitals and offices throughout the sample period. If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction. As this type of hospitalist is less sticky, if there is any effect of EHR exposure on work setting it would be among this type.

@@

There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation. A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment. For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment. An important aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.

Whether EHRs affect productivity in terms of patients seen is theoretically ambiguous. If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients. In this case, I expect productivity to increase. However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.

Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the physician. Billing activity refers to the number of claims filed per patient seen. EHRs may improve efficiency if they remove the need for repeated or excess testing. However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient. Thus, both productivity and efficiency are empirical questions.

@@

There is still a large portion of physicians not induced to retire due to EHR exposure. Thus, I consider the decision to switch work settings using multiple outcomes. First, I investigate the extent to which hospitalists work in an office and whether this changes due to EHR exposure. I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office in a given year. Second, I investigate whether EHR exposure causes a change in the likelihood of switching zip codes. This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.

I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}. The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the same year as exposure by .014 ppts, a 4.6\% increase relative to the mean. This result indicates that hospitalists are more likely to start working in an office in the year they are exposed to an EHR, though the effect is not as large as the one found for physicians retiring. Splitting the sample by age reveals that this result is driven by the younger sample of physicians. For retirement age physicians, the estimates become very noisy and I do not draw any conclusions. Younger physicians, however, show evidence of switching behavior. The .015 ppt increase in the likelihood of retiring is equivalent to a 5.4\% increase relative to the mean for this age group. While there is not strong evidence for the existence of a pre-trend, I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}. This analysis yields the same result.

@@

The previous results is driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians shift the amount of work they do in an office. Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is also tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office, and the results are the same.} I show the estimates and 95\% confidence intervals for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings. This does not seem to be a margin where physicians are changing their behavior due to EHR use.

\begin{figure}[ht]
    \centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
    \includegraphics[scale=.6]{Objects/officefrac__plot.pdf}
    \label{fig:officesecond}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0003 (0.004), -0.0007 (0.005), -0.006 (0.018), respectively.}}
\end{figure}

The previous results only capture switching behavior which involves office based settings. Another way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors. Again, I exclude physicians who ever retire. The outcome variable is equal to 1 if their list of primary zip codes changes. This variable is limited in how broadly it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes. While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation that can be combined with the previous outcome variables to draw conclusions about general behavior.

The effects of EHR exposure on changing zip codes over time are shown in Figure \ref{fig:zip}. For the full sample of hospitalists, there is a statistically positive effect observed in each year after EHR exposure, but there is evidence that this result is driven by a pre-trend before exposure. Therefore, I do not make any causal arguments about this effect. However, I include a plot of confidence intervals that are robust to specified violations of parallel trends for each age group in Appendix \ref{sec:pretrends}. These results reveal that under linear violations of the parallel trends assumption, there is still a positive relationship between EHR exposure and changing zip codes for younger physicians. When taken together, these estimates suggest there is slight evidence of physicians who do not retire changing aspects of their work environment, but not on a large scale, indicating stickiness in place of work.

@@

\subsection{Patient Count and Billing Activity}\label{sec:patientcount}

As a measure of productivity or efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient. If the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the change in environment, not due to the EHR. Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample. These hospitalists do not retire or see patients in another hospital. Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the sample.

The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}. For hospitalists of any age, patient count increases by 17.5 patients in the year of EHR exposure, 34.7 patients in the year after exposure, 16.6 patients in the second year after exposure, and then the effect disappears. These estimates translate to a 5-10\% increase in the number of patients seen relative to the mean. Interestingly, when splitting the sample by age group, the effect becomes null for those of retirement age, suggesting that technology use impacts productivity dependent on worker type (in this case, age). Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends}.

@@

While this paper is interested the supply side factors of patients seen, this variable is also driven by demand for health care. A major change in the U.S. health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook. If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count. The ACA had potentially opposing effects on Medicare patients: Medicare patients may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population. A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results. Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}. Each group shows the same pattern of a small increase in patient count in the first 0-2 years after implementation.

@@

Since I found in the earlier stages of analysis that physicians may have shifted away from hospitals due to EHR exposure, I also investigate whether this increase in the number of patients seen is due to the decrease in working physicians. For example, if a hospital loses a portion of their physicians, they may divide those patients between the remaining physicians. In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away then these estimates could be driven by the retiring physicians instead of EHR use. In Figure \ref{fig:numphys}, I present a graph showing the relationship between EHR adoption and the number of physicians working in the hospital. While these are not causal estimates, they do show that the trend seems to be more physicians working in hospitals, and there is no apparent shock in the short run after EHR implementation. This suggests that the increase in patient count is not due to the exit of physicians observed in the earlier analysis.

@@

Next, I consider whether EHR exposure affects claims filed per patient. On average, claim counts are roughly 4 times the number of patients seen in a year. Shown in Figure \ref{fig:claim}, there is a decreasing trend in claims per patient in the three years after EHR exposure. Two years after exposure, claims per patient decreases by .25, roughly 6\% of the average. This indicates a potentially advantageous effect of EHRs, that excessive billing is reduced. Though noisier for retirement age physicians, this effect is seen in both age groups.

\begin{figure}[ht]
    \centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Claims per Patient}
    \includegraphics[scale=.6]{Objects/claim_per_patient_plot.pdf}
    \label{fig:claim}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 4.04 (33.89), 25.77 (31.26), -161.06 (133.35), respectively.}}
\end{figure}

@@

For the results presented, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure. Almost all specifications point to the effect of EHR exposure becmoing null in the 3-4 years after exposure. One reason for this may be the composition of the hospitals which adopt at different times. Consider event time $t+4$. The physician included in the estimation for this coefficient are those exposed in 2010 compared with those who did not adopt until 2015, implying that the underlying hospitals implemented EHRs in either 2010 or 2015. It is likely that the group of hospitals who adopted in 2015 are compositionally different than early adopters, especially if poorer hospitals adopted later. For this reason, I have focused discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.

@@

The effects found in the above analysis could be biased if there is an external event occurring that is correlated with EHR implementation and exposure. The existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology. Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals. they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator. I will refer to any employee in these categories as a data assistant. In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs. This is particularly relevant for the productivity and efficiency outcomes where I find that EHRs improve both outcomes.

@@

In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes. This technology has been controversial and debated as a burdensome load on physicians that may contribute to physician burnout, and therefore patient outcomes. Understanding how physicians respond to this technology is important as physician labor markets have implications for access to care, quality of care, and costs.

I utilize data on shared patients, hospital EHR use, and physician billing activity. Using a difference in difference framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR, where those of retirement age have an even larger increase relative to younger physicians. Further, I find slight evidence that physicians who do not retire change where they work because of EHR exposure. Finally, physicians who utilize EHRs exhibit an increase in the number of patients seen and a decrease in claims filed per patient.

The implications of these results are threefold. First, a surge of physicians left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening access to care. In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss of senior physicians may have long term negative effects on patient outcomes, but this is a question left to future research. Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens. When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation. Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users. While the technology may be a burden, physicians are able to see more patients because of their use, and do not seem to over-bill because of them.

@@

For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}}. I use data that captures shared patient activity in 30 day intervals. I merge the shared patient data to the filtered tax code data created in Section \ref{sec:taxcode} to identify NPIs who are either PCPs or hospitals. I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs. Some pairs are duplicates from the hospital being listed first or second in the shared patient data. I combine duplicates into one observation, summing the same day count variable. Once duplicated are removed, there are 7.1 million observations. Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital. I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60. Results are similar and can be found in Figure ____.} The final list of pairs consists of 1.3 million observations.

@@
